\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-28 tex/paper.tex
save

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

For continuous data, we first consider inference for
a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.8cm}
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
             \sigma|\psi)$ over the generator parameters.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
\end{figure}
\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates the effectiveness of adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
\end{figure}

@@

Similarly, we evaluate adversarial variational optimization on discrete data,
for the inference of the scale parameter $\lambda$ of a Poisson distribution.
The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
initialized with a mean at $\log 5$ and unit variance. The optimization setup is
otherwise left unchanged. Despite the discreteness and the non-differentiability
of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
adversarial variational optimization works. The upper plot in the figure shows
that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
the proposal distribution concentrated at $\log 7 = 1.94$.

\subsection{Plinko}

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}

>>>>

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

{\it Continuous data.}
We first consider inference for
a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.8cm}
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
             \sigma|\psi)$ over the generator parameters.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
\end{figure}
\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
\end{figure}

@@

{\it Discrete data.}
Similarly, we evaluate adversarial variational optimization on discrete data,
for the inference of the scale parameter $\lambda$ of a Poisson distribution.
The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
initialized with a mean at $\log 5$ and unit variance. The optimization setup is
otherwise the same as for the continuous case. Despite the discreteness and the non-differentiability
of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
adversarial variational optimization works. The upper plot in the figure shows
that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
the proposal distribution concentrated at $\log 7 = 1.94$.

\subsection{Galton box}

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}
\glnote{next version = try on an actual particle physics simulator.}

\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}
